They're Clutching up! Team Momentum in Round-Based Esports

### 2022-08-02 presentation.qmd
formalize streaking with some FUNctions, basic plotting code

## Is there evidence for the "hot hand"?

@@

::: fragment
Looking at streaks of 3
:::

```{r}
#| label: cod_round_win_prop_after_b2b2b_wins
## http://keyonvafa.com/hot-hand/
get_post_streak_prob <- function(n, k, p = 0.5) {
  tosses <- rbinom(n, 1, p)
  runs <- rle(tosses)
  n_neg_after <- length(which(runs$values == 1 & runs$lengths >= k))
  n_pos_after <- sum(runs$lengths[which(runs$values == 1 & runs$lengths >= k)] - k)

@@

summarize_cod_win_prop_after_x_wins <- function(kind = c('b2b', 'b2b2b')) {
  kind <- match.arg(kind)
  is_b2b <- kind == 'b2b'
  if (is_b2b) {
    k <- 2
  } else {
    k <- 3
  }

  probs_nx_k <- tibble(
    n_rounds = 7L:11L
  ) |> 
    mutate(
      trial_p = 6 / n_rounds,
      expected_poststreak_p = map2_dbl(
        n_rounds, trial_p, 
        ~simulate_post_streak_prob(n = ..1, k = !!k, p = ..2)
      )
    )

  cod_round_win_prop_after_x_wins <- if (is_b2b) {
    cod_round_streaks_after_x <- cod_round_streaks |> 
      filter(!is.na(won_prior_round2)) |> 
      count(n_rounds, win_round, won_prior_round, won_prior_round2, sort = TRUE)

     cod_round_streaks_after_x |> 
      filter(won_prior_round, won_prior_round2) 
  } else {
    cod_round_streaks_after_x <- cod_round_streaks |> 
      filter(!is.na(won_prior_round2), !is.na(won_prior_round3)) |> 
      count(n_rounds, win_round, won_prior_round, won_prior_round2, won_prior_round3, sort = TRUE)

     cod_round_streaks_after_x |> 
      filter(won_prior_round, won_prior_round2, won_prior_round3) 
  }

@@

cod_round_win_prop_after_b2b_wins <- summarize_cod_win_prop_after_x_wins('b2b')
cod_round_win_prop_after_b2b2b_wins <- summarize_cod_win_prop_after_x_wins('b2b2b')
```

```{r}
#| label: cod_round_win_prop_after_b2b2b_wins-eval
#| eval: true
# cod_round_win_prop_after_b2b_wins |> clipr::write_clip() |> datapasta::tribble_paste()
cod_round_win_prop_after_b2b_wins <- tibble::tribble(
  ~n_rounds,   ~n, ~total,             ~prop,          ~trial_p, ~expected_poststreak_p,       ~gt_p_value,          ~lt_p_value,
         7L, 262L,   328L, 0.798780487804878, 0.857142857142857,      0.786573045059608,   0.3220219536276,    0.725316732579212,
         8L, 265L,   421L, 0.629453681710214,              0.75,      0.648791374811152, 0.811481245561487,    0.217078977854197,
         9L, 275L,   577L, 0.476603119584055, 0.666666666666667,      0.557203226372085, 0.999956606998609, 6.11343021400038e-05,
        10L, 257L,   628L, 0.409235668789809,               0.6,      0.483529268542724, 0.999920968231117, 0.000108871179188732,
        11L, 317L,   807L, 0.392812887236679, 0.545454545454545,       0.42596043311709, 0.974133848646825,   0.0304921375685209
  )
# cod_round_win_prop_after_b2b2b_wins |> clipr::write_clip() |> datapasta::tribble_paste()
cod_round_win_prop_after_b2b2b_wins <- tibble::tribble(
  ~n_rounds,   ~n, ~total,             ~prop,          ~trial_p, ~expected_poststreak_p,       ~gt_p_value,         ~lt_p_value,
         7L, 149L,   201L, 0.741293532338308, 0.857142857142857,      0.757307821042062, 0.732554638281868,   0.323308922871945,
         8L, 126L,   203L, 0.620689655172414,              0.75,      0.618543392752532, 0.505988058731931,   0.551527754692743,
         9L,  98L,   209L,   0.4688995215311, 0.666666666666667,      0.526500674199098, 0.958718294702725,   0.055063286515911,
        10L,  74L,   208L, 0.355769230769231,               0.6,      0.446500557074851, 0.996786203815779, 0.00490308277112825,
        11L,  90L,   276L, 0.326086956521739, 0.545454545454545,      0.387843835660101, 0.985650077371605,  0.0196488641002811
  )
```

::: fragment
There is no evidence for momentum, for both streaks of 2 and 3. In fact, for series that go to 9, 10, or 11 rounds, there is evidence that the streaking teams perform significantly worse than expected in the post-streak round.
:::

>>>>

## Is there evidence for the "hot hand" ([COD SnD]{.cod})?

@@

```{r}
#| label: cod_round_win_prop_after_x_wins
## http://keyonvafa.com/hot-hand/
get_post_streak_prob <- function(n, k, p = 0.5) {
  tosses <- rbinom(n, 1, p)
  runs <- rle(tosses)
  n_neg_after <- length(which(runs$values == 1 & runs$lengths >= k))
  n_pos_after <- sum(runs$lengths[which(runs$values == 1 & runs$lengths >= k)] - k)

@@

cod_round_streaks <- cod_rounds |> 
  group_by(series_id, team) |> 
  mutate(
    won_prior_round1 = lag(win_round, n = 1, default = NA),
    won_prior_round2 = lag(win_round, n = 2, default = NA),
    won_prior_round3 = lag(win_round, n = 3, default = NA),
    won_prior_round4 = lag(win_round, n = 4, default = NA),
    won_prior_round5 = lag(win_round, n = 5, default = NA)
  ) |> 
  ungroup() |>
  select(
    series_id,
    team,
    is_offense,
    round,
    n_rounds,
    win_round,
    starts_with('won_prior_round')
  )

summarize_cod_win_prop_after_x_wins <- function(k) {
  stopifnot((k >= 2) & (k <= 5))

  probs_nx_k <- tibble(
    n_rounds = 7L:11L
  ) |> 
    mutate(
      prior_p = 6L / n_rounds,
      expected_poststreak_p = map2_dbl(
        n_rounds, prior_p, 
        ~simulate_post_streak_prob(n = ..1, k = !!k, p = ..2)
      )
    )

  cols <- sprintf('won_prior_round%d', 2:k)
  col_syms <- syms(cols)
  cod_round_streaks_after_x <- cod_round_streaks |> 
    drop_na(!!!col_syms) |> 
    count(n_rounds, win_round, won_prior_round1, !!!col_syms, sort = TRUE)

  cod_round_win_prop_after_x_wins <- cod_round_streaks_after_x |> 
    filter(won_prior_round1, !!!col_syms)

@@

cod_round_win_prop_after_x_wins <- tibble(win_streak = 2L:4L) |> 
  mutate(
    data = map(win_streak, summarize_cod_win_prop_after_x_wins)
  ) |> 
  unnest(data)
```

```{r}
#| label: cod_round_win_prop_after_x_wins-eval
#| eval: true
cod_round_win_prop_after_x_wins |> 
  transmute(
    `Win Streak` = win_streak,
    `Rounds` = n_rounds,
    `N` = scales::comma(total),
    across(c(prop, expected_poststreak_p), ~scales::percent(.x, accuracy = 0.1)),
    across(
      ends_with('p_value'), 
      ~case_when(
        .x <= 0.01 ~ '<=0.01', 
        round(.x, 2) == 1 ~ '1',
        TRUE ~ as.character(round(.x, 2))
      )
    )
  ) |>
  rename(
    `Actual post-streak win %` = prop,
    `Expected post-streak win %` = expected_poststreak_p,
    `Binom. > p-value` = gt_p_value,
    `Binom. < p-value` = lt_p_value
  ) |> 
  knitr::kable()

::: fragment
-   There is no evidence for momentum, for both streaks of 2 and 3. In fact, for series that go to 9, 10, or 11 rounds, there is evidence that the streaking teams perform significantly worse than expected in the post-streak round. And for streaks of 4, teams perform worse than expected in series ending in 7, 8, 10, and 11 rounds.
:::

## Is there evidence for the "hot hand" ([Valorant]{.valorant})?

```{r}
#| label: val_round_win_prop_after_x_wins-save
#| eval: false
val_round_streaks <- val_rounds |> 
  group_by(series_id, team, is_offense) |> 
  mutate(
    won_prior_round1 = lag(win_round, n = 1, default = NA),
    won_prior_round2 = lag(win_round, n = 2, default = NA),
    won_prior_round3 = lag(win_round, n = 3, default = NA),
    won_prior_round4 = lag(win_round, n = 4, default = NA),
    won_prior_round5 = lag(win_round, n = 5, default = NA),
    won_prior_round6 = lag(win_round, n = 6, default = NA),
    won_prior_round7 = lag(win_round, n = 7, default = NA),
    won_prior_round8 = lag(win_round, n = 8, default = NA),
    won_prior_round9 = lag(win_round, n = 9, default = NA)
  ) |> 
  ungroup() |>
  select(
    series_id,
    team,
    is_offense,
    cumu_w,
    round,
    n_rounds,
    win_round,
    starts_with('won_prior_round')
  )

summarize_val_win_prop_after_x_wins_first12r <- function(k) {
  stopifnot((k >= 3) & (k <= 9))
  probs_nx_k <- tibble(
    cumu_w = 4L:11L,
    n_rounds = 12L
  ) |> 
    mutate(
      prior_p = cumu_w / n_rounds,
      expected_poststreak_p = map2_dbl(
        n_rounds, prior_p, 
        ~simulate_post_streak_prob(n = ..1, k = !!k, p = ..2)
      )
    )

  cols <- sprintf('won_prior_round%d', 2:k)
  col_syms <- syms(cols)
  val_round_streaks_after_x <- val_round_streaks |> 
    filter(round < 13) |> 
    drop_na(!!!col_syms) |> 
    count(cumu_w, win_round, won_prior_round1, !!!col_syms, sort = TRUE)

  val_round_win_prop_after_x_wins <- val_round_streaks_after_x |> 
    filter(won_prior_round1, !!!col_syms)

  val_round_win_prop_after_x_wins |> 
    group_by(cumu_w) |> 
    mutate(
      total = sum(n),
      prop = n / total
    ) |> 
    ungroup() |> 
    filter(win_round) |> 
    select(cumu_w, n, total, prop) |> 
    arrange(cumu_w) |> 
    inner_join(
      probs_nx_k,
      by = 'cumu_w'
    ) |> 
    nest(data = -c(cumu_w)) |> 
    mutate(
      gt_p_value = map_dbl(
        data, 
        ~binom.test(.x$n, .x$total, .x$expected_poststreak_p, alternative = 'greater') |> 
          broom::tidy() |>
          pull(p.value) |> 
          round(3)
      ),
      lt_p_value = map_dbl(
        data, 
        ~binom.test(.x$n, .x$total, .x$expected_poststreak_p, alternative = 'less') |> 
          broom::tidy() |>
          pull(p.value) |> 
          round(3)
      )
    ) |> 
    unnest(data)
}

summarize_val_win_prop_after_x_wins_second12r <- function(k) {
  stopifnot((k >= 3) & (k <= 9))
  probs_nx_k <- crossing(
    cumu_w = 0L:13L,
    n_rounds = 13L:24L
  ) |> 
    mutate(
      prior_p = cumu_w / n_rounds,
      expected_poststreak_p = map2_dbl(
        n_rounds, prior_p, 
        ~simulate_post_streak_prob(n = ..1, k = !!k, p = ..2)
      )
    )

  cols <- sprintf('won_prior_round%d', 2:k)
  col_syms <- syms(cols)
  val_round_streaks_after_x <- val_round_streaks |> 
    filter(round > 12) |> 
    drop_na(!!!col_syms) |> 
    count(n_rounds, cumu_w, win_round, won_prior_round1, !!!col_syms, sort = TRUE)

  val_round_win_prop_after_x_wins <- val_round_streaks_after_x |> 
    filter(won_prior_round1, !!!col_syms)

  val_round_win_prop_after_x_wins |> 
    group_by(n_rounds, cumu_w) |> 
    mutate(
      total = sum(n),
      prop = n / total
    ) |> 
    ungroup() |> 
    filter(win_round) |> 
    select(n_rounds, cumu_w, n, total, prop) |> 
    arrange(n_rounds, cumu_w) |> 
    inner_join(
      probs_nx_k,
      by = c('n_rounds', 'cumu_w')
    ) |> 
    nest(data = -c(n_rounds, cumu_w)) |> 
    mutate(
      gt_p_value = map_dbl(
        data, 
        ~binom.test(.x$n, .x$total, .x$expected_poststreak_p, alternative = 'greater') |> 
          broom::tidy() |>
          pull(p.value) |> 
          round(3)
      ),
      lt_p_value = map_dbl(
        data, 
        ~binom.test(.x$n, .x$total, .x$expected_poststreak_p, alternative = 'less') |> 
          broom::tidy() |>
          pull(p.value) |> 
          round(3)
      )
    ) |> 
    unnest(data)
}

save_qs <- function(df, name = deparse(substitute(df))) {
  qs::qsave(df, sprintf('data/%s.qs', name))
}

val_round_win_prop_after_x_wins_first12r <- tibble(win_streak = 3L:9L) |> 
  mutate(
    data = map(win_streak, summarize_val_win_prop_after_x_wins_first12r)
  ) |> 
  unnest(data)
save_qs(val_round_win_prop_after_x_wins_first12r)

val_round_win_prop_after_x_wins_second12r <- tibble(win_streak = 3L:9L) |> 
  mutate(
    data = map(win_streak, summarize_val_win_prop_after_x_wins_second12r)
  ) |> 
  unnest(data)
save_qs(val_round_win_prop_after_x_wins_second12r)
```

```{r}
#| label: val_round_win_prop_after_x_wins
#| eval: false
load_qs <- function(name) {
  res <- qs::qread(sprintf('data/%s.qs', name))
  assign(name, value = res, envir = .GlobalEnv)
}
load_qs('val_round_win_prop_after_x_wins_first12r')
load_qs('val_round_win_prop_after_x_wins_second12r')
val_round_win_prop_after_x_wins_first12r |> 
  filter(n > 30) |> 
  mutate(
    gt_signif = gt_p_value <= 0.05
  ) |> 
  ggplot() +
  aes(x = win_streak, y = cumu_w) +
  geom_tile(
    aes(fill = gt_signif)
  ) +
  geom_text(
    aes(label = sprintf(
      '%s | %s\n(%s)',
      scales::percent(prop, accuracy = 1),
      scales::percent(expected_poststreak_p, accuracy = 1),
      scales::comma(total)
      )
    )
  )

val_round_win_prop_after_x_wins_second12r |> 
  filter(n > 30) |> 
  mutate(
    gt_signif = gt_p_value <= 0.05
  ) |> 
  ggplot() +
  aes(x = win_streak, y = cumu_w) +
  geom_tile(
    aes(fill = gt_signif)
  ) +
  geom_text(
    aes(label = sprintf(
      '%s | %s\n(%s)',
      scales::percent(prop, accuracy = 1),
      scales::percent(expected_poststreak_p, accuracy = 1),
      scales::comma(total)
      )
    )
  ) +
  facet_wrap(~n_rounds)

```{r}
#| label: cod_round_win_prop_after_xw-eval
#| eval: true
val_round_win_prop_after_xw |> 
  transmute(
    `Win Streak` = win_streak,
    `# of Offensive Wins in First 12 Rounds` = cumu_w,
    `N` = scales::comma(total),
    across(c(prop, expected_poststreak_p), ~scales::percent(.x, accuracy = 0.1)),
    across(
      ends_with('p_value'), 
      ~case_when(
        .x <= 0.01 ~ '<=0.01', 
        round(.x, 2) == 1 ~ '1',
        TRUE ~ as.character(round(.x, 2))
      )
    )
  ) |>
  rename(
    `Actual post-streak win %` = prop,
    `Expected post-streak win %` = expected_poststreak_p,
    `Binom. > p-value` = gt_p_value,
    `Binom. < p-value` = lt_p_value
  ) |> 
  knitr::kable()

::: fragment
First 12 rounds **only**
:::

::: fragment
-   There is no evidence for momentum, for both streaks of 2 and 3. In fact, for series that go to 9, 10, or 11 rounds, there is evidence that the streaking teams perform significantly worse than expected in the post-streak round. And for streaks of 4, teams perform worse than expected in series ending in 7, 8, 10, and 11 rounds.
:::

::: fragment
-   Given the nature of how economy is used in the game, perhaps this is not suprising at all.
:::
